\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-08-12 Papers/Stochastic_Uncertainty_Propagation_TPS/JKO_Power_Systems_TPS_Rev2.tex
Rebuttal round 2

\begin{remark}\label{remark:timevaryingEffectivePowerInput}
The proposed computational framework is also applicable when $P_{i}^{\rm{mech}}, P_{i}^{\rm{load}}$, and thus $P_{i}$ in \eqref{defPi}, are bounded time-varying functions. This can be the case, for instance, with rapid fluctuations from renewables {\blue{or from loads}} in the short term.
\end{remark}

@@

\noindent\textbf{Case I.} \textbf{Nominal case} where the $P_{i}^{\text{mech}}$ and $P_{i}^{\text{load}}$ are from the steady state power flow solutions without the process noise, and {\blue{are assumed}} to remain constant over the simulation time horizon. The transient stochastic simulations (i.e., with process noise) are then performed using the proposed framework.

@@

With step size $h=10^{-3}$, we discretized time $t\in[0,1]$. Using the algorithm detailed in Sec. \ref{sec:ProxAlgo}, for both cases, we propagated $N=1000$ random {\blue{state}} samples from \eqref{rho0IEEE14}, and {\blue{corresponding joint PDF values}} evaluated at \eqref{rho0IEEE14}, to update the scattered weighted point clouds $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ over $\mathbb{T}^{5}\times\mathbb{R}^{5}$ at times $t_{k}:=kh$, $k\in\mathbb{N}$. In Algorithm \ref{algo:KineticProx}, we used the algorithmic parameters $\varepsilon=0.05$, $\delta=10^{-3}$, $\ell_{\max}=300$.

@@

Fig. \ref{fig:IEEE14omegamarginalsCase12} shows the corresponding univariate $\omega_{i}$-marginal PDFs. In particular, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) helps compare the $\omega_{4}$ marginal evolution for bus 6 in the nominal (\textbf{Case I}) and post-fault (\textbf{Case II}) scenarios. Since \textbf{Case II} considers the failure of line 13, from Fig. \ref{fig:Motivating} we intuitively expect that the transient effects of this fault will be prominent for bus 6 (generator 4). Indeed, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) shows that although the $\omega_{4}$ marginals in both \textbf{Case I} and \textbf{II}, at all times have the highest probability around $0$ rad/s, the post-fault marginals show larger spread or dispersion of the probability mass, as expected. That the generator 4 states in \textbf{Case II} have larger uncertainties than in \textbf{Case I}, is also visible in the bivariate marginal plots if we compare the fourth row in Fig. \ref{fig:IEEE14marginalsCase0} with the same in Fig. \ref{fig:IEEE14marginalsCase1}. {\blue{The statistical uncertainties in the rotor angles are shown in Fig. \ref{fig:IEEE14theta} for both \textbf{Case I} and \textbf{Case II}.}}

Fig. \ref{fig:CompTimeIEEE14bus} shows the computational times needed to update $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ via the algorithm in Sec. \ref{sec:ProxAlgo}, for both \textbf{Cases I} and \textbf{II}. {\blue{While the depiction of the computational times in Fig. \ref{fig:CompTimeIEEE14bus} appears unconventional, we next explain why that is meaningful in our framework.

@@

Fig. \ref{fig:TotalCompTime} (circle markers for the \textbf{Case I}, diamond markers for the \textbf{Case II}) shows the total computational times needed for propagating the transient joint state PDFs over a longer horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.
}}

@@

Fig. \ref{fig:CompTimeSynthetic} highlights that for computing high dimensional transient joint state PDFs as in here, the proposed variational framework enjoys remarkably fast computational time. Notice that constructing histograms or other conventional density estimators for the joint state PDF by making a grid over the 100 dimensional state space, is computationally prohibitive in a general purpose computer. {\blue{Fig. \ref{fig:TotalCompTime} (square markers) shows the total computational times needed for propagating the transient joint state PDFs over the time horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.}}

To depict the numerical accuracy, Fig. \ref{fig:RelErrSynthetic} plots the (time-varying) relative error between the empirical (i.e., Monte Carlo estimate) mean vector $\bm{\mu}_{k}^{\text{MC}}\in\mathbb{T}^{50}\times\mathbb{R}^{50}$, and the ``proximal mean" vector $\bm{\mu}_{k}^{\text{Prox}}\in\mathbb{T}^{50}\times\mathbb{R}^{50}$. The latter was computed using the proximal updates $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$. {\blue{Likewise, Fig. \ref{fig:RelErrCovSynthetic} plots the time-varying normalized Bures-Wasserstein distance $d_{\rm{BW}}(\cdot,\cdot)$ between the empirical covariance matrix (i.e., Monte Carlo estimate) ${\rm{Cov}}^{\rm{MC}}\in\mathbb{R}^{100\times 100}$, and the ensemble covariance matrix ${\rm{Cov}}^{\rm{Prox}}\in\mathbb{R}^{100\times 100}$ obtained using the proximal updates. The unnormalized Bures-Wasserstein distance \cite{bhatia2019bures} given by 
\begin{align*}
&d_{\rm{BW}}(\rm{Cov}^{\rm{MC}},\rm{Cov}^{\rm{Prox}}) := \left[\operatorname{trace}\left(\rm{Cov}^{\rm{MC}}+\rm{Cov}^{\rm{Prox}}\right)\right.\\
&\qquad\qquad \left.-2\operatorname{trace}\left(\left(\sqrt{\rm{Cov}^{\rm{MC}}}\:\rm{Cov}^{\rm{Prox}}\:\sqrt{\rm{Cov}^{\rm{MC}}}\right)^{\frac{1}{2}}\right)\right]^{\frac{1}{2}},	
\end{align*}
is a metric on the cone of symmetric positive definite matrices such as the covariance matrices, and can be seen as an absolute error between the covariances. Thus, Fig. \ref{fig:RelErrCovSynthetic} can be viewed as a relative error curve between the covariances. In summary, Figs. \ref{fig:RelErrSynthetic} and \ref{fig:RelErrCovSynthetic} show that the time-varying statistics between the Monte Carlo and the proposed proximal computation remain close at all times.}}

{\blue{Since it is not possible to \emph{a priori} determine the dimension of the sufficient statistic of the transient joint state PDFs, Fig. \ref{fig:WassJointPDFSynthetic} shows the 2-Wasserstein metric between the empirical joint PDF (i.e., Monte Carlo estimate) $\rho^{\rm{MC}}$, and the proximal update $\varrho^{k}$. In other words, Fig. \ref{fig:WassJointPDFSynthetic} can be viewed as a nonparametric absolute error curve between $\rho^{\rm{MC}}$ and $\varrho^{k}$. Each snapshot value in Fig. \ref{fig:WassJointPDFSynthetic} is computed by solving the well-known optimal transport linear program (see e.g., \cite[Sec. 4.2-4.3]{halder2014probabilistic}, \cite[Ch. 3]{peyre2019computational}) using the corresponding time-varying weighted point clouds.}}

@@

{\blue{\bibitem{wang2011numerical}
K.~Wang, and M.L.~Crow, ``Numerical simulation of stochastic differential algebraic equations for power system transient stability with random loads", \emph{2011 IEEE Power \& Energy Society General Meeting (PESGM)}, pp. 1--8, 2011. doi: 10.1109/PES.2011.6039188

\bibitem{milano2013systematic}
F.~Milano, and R.~Z{\'a}rate-Mi{\~n}ano, ``A systematic method to model power systems as stochastic differential algebraic equations", \emph{IEEE Transactions on Power Systems}, Vol. 28, No. 4, pp. 4537--4544, 2013.}}

@@

{\blue{\bibitem{geng2020reach}
S.~Geng, and I.A.~Hiskens, ``Reach-Set Estimation for DAE Systems under Uncertainty and Disturbances Using Trajectory Sensitivity and Logarithmic Norm", \emph{21st IFAC World Congress}, Vol. 53, No. 2, pp. 1955--1961, 2020.}}

@@

\bibitem{silverman1998density}
{\blue{B.W.~Silverman, \emph{Density Estimation for Statistics and Data Analysis}, 1st ed., Chapman \& Hall/CRC, 1998.}}

@@

{\blue{\bibitem{bhatia2019bures}
R.~Bhatia, T.~Jain, and Y.~Lim, ``On the Bures--Wasserstein Distance between Positive Definite Matrices", \emph{Expositiones Mathematicae}, Vol. 37, No. 2, pp. 165--191, 2019.

@@

}}

>>>>

\begin{remark}\label{remark:timevaryingEffectivePowerInput}
The proposed computational framework is also applicable when $P_{i}^{\rm{mech}}, P_{i}^{\rm{load}}$, and thus $P_{i}$ in \eqref{defPi}, are bounded time-varying functions. This can be the case, for instance, with rapid fluctuations from renewables or from loads in the short term.
\end{remark}

@@

\noindent\textbf{Case I.} \textbf{Nominal case} where the $P_{i}^{\text{mech}}$ and $P_{i}^{\text{load}}$ are from the steady state power flow solutions without the process noise, and are assumed to remain constant over the simulation time horizon. The transient stochastic simulations (i.e., with process noise) are then performed using the proposed framework.

@@

With step size $h=10^{-3}$, we discretized time $t\in[0,1]$. Using the algorithm detailed in Sec. \ref{sec:ProxAlgo}, for both cases, we propagated $N=1000$ random state samples from \eqref{rho0IEEE14}, and corresponding joint PDF values evaluated at \eqref{rho0IEEE14}, to update the scattered weighted point clouds $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ over $\mathbb{T}^{5}\times\mathbb{R}^{5}$ at times $t_{k}:=kh$, $k\in\mathbb{N}$. In Algorithm \ref{algo:KineticProx}, we used the algorithmic parameters $\varepsilon=0.05$, $\delta=10^{-3}$, $\ell_{\max}=300$.

@@

Fig. \ref{fig:IEEE14omegamarginalsCase12} shows the corresponding univariate $\omega_{i}$-marginal PDFs. In particular, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) helps compare the $\omega_{4}$ marginal evolution for bus 6 in the nominal (\textbf{Case I}) and post-fault (\textbf{Case II}) scenarios. Since \textbf{Case II} considers the failure of line 13, from Fig. \ref{fig:Motivating} we intuitively expect that the transient effects of this fault will be prominent for bus 6 (generator 4). Indeed, Fig. \ref{fig:IEEE14omegamarginalsCase12}(c) shows that although the $\omega_{4}$ marginals in both \textbf{Case I} and \textbf{II}, at all times have the highest probability around $0$ rad/s, the post-fault marginals show larger spread or dispersion of the probability mass, as expected. That the generator 4 states in \textbf{Case II} have larger uncertainties than in \textbf{Case I}, is also visible in the bivariate marginal plots if we compare the fourth row in Fig. \ref{fig:IEEE14marginalsCase0} with the same in Fig. \ref{fig:IEEE14marginalsCase1}. The statistical uncertainties in the rotor angles are shown in Fig. \ref{fig:IEEE14theta} for both \textbf{Case I} and \textbf{Case II}.

Fig. \ref{fig:CompTimeIEEE14bus} shows the computational times needed to update $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$ via the algorithm in Sec. \ref{sec:ProxAlgo}, for both \textbf{Cases I} and \textbf{II}. While the depiction of the computational times in Fig. \ref{fig:CompTimeIEEE14bus} appears unconventional, we next explain why that is meaningful in our framework.

@@

Fig. \ref{fig:TotalCompTime} (circle markers for the \textbf{Case I}, diamond markers for the \textbf{Case II}) shows the total computational times needed for propagating the transient joint state PDFs over a longer horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.

@@

Fig. \ref{fig:CompTimeSynthetic} highlights that for computing high dimensional transient joint state PDFs as in here, the proposed variational framework enjoys remarkably fast computational time. Notice that constructing histograms or other conventional density estimators for the joint state PDF by making a grid over the 100 dimensional state space, is computationally prohibitive in a general purpose computer. Fig. \ref{fig:TotalCompTime} (square markers) shows the total computational times needed for propagating the transient joint state PDFs over the time horizon $[0, 1\;\text{min}]$ for 100 different simulation instances.

To depict the numerical accuracy, Fig. \ref{fig:RelErrSynthetic} plots the (time-varying) relative error between the empirical (i.e., Monte Carlo estimate) mean vector $\bm{\mu}_{k}^{\text{MC}}\in\mathbb{T}^{50}\times\mathbb{R}^{50}$, and the ``proximal mean" vector $\bm{\mu}_{k}^{\text{Prox}}\in\mathbb{T}^{50}\times\mathbb{R}^{50}$. The latter was computed using the proximal updates $\{\bm{x}_{k}^{i},\varrho_{k}^{i}\}_{i=1}^{N}$. Likewise, Fig. \ref{fig:RelErrCovSynthetic} plots the time-varying normalized Bures-Wasserstein distance $d_{\rm{BW}}(\cdot,\cdot)$ between the empirical covariance matrix (i.e., Monte Carlo estimate) ${\rm{Cov}}^{\rm{MC}}\in\mathbb{R}^{100\times 100}$, and the ensemble covariance matrix ${\rm{Cov}}^{\rm{Prox}}\in\mathbb{R}^{100\times 100}$ obtained using the proximal updates. The unnormalized Bures-Wasserstein distance \cite{bhatia2019bures} given by 
\begin{align*}
&d_{\rm{BW}}(\rm{Cov}^{\rm{MC}},\rm{Cov}^{\rm{Prox}}) := \left[\operatorname{trace}\left(\rm{Cov}^{\rm{MC}}+\rm{Cov}^{\rm{Prox}}\right)\right.\\
&\qquad\qquad \left.-2\operatorname{trace}\left(\left(\sqrt{\rm{Cov}^{\rm{MC}}}\:\rm{Cov}^{\rm{Prox}}\:\sqrt{\rm{Cov}^{\rm{MC}}}\right)^{\frac{1}{2}}\right)\right]^{\frac{1}{2}},	
\end{align*}
is a metric on the cone of symmetric positive definite matrices such as the covariance matrices, and can be seen as an absolute error between the covariances. Thus, Fig. \ref{fig:RelErrCovSynthetic} can be viewed as a relative error curve between the covariances. In summary, Figs. \ref{fig:RelErrSynthetic} and \ref{fig:RelErrCovSynthetic} show that the time-varying statistics between the Monte Carlo and the proposed proximal computation remain close at all times.

Since it is not possible to \emph{a priori} determine the dimension of the sufficient statistic of the transient joint state PDFs, Fig. \ref{fig:WassJointPDFSynthetic} shows the 2-Wasserstein metric between the empirical joint PDF (i.e., Monte Carlo estimate) $\rho^{\rm{MC}}$, and the proximal update $\varrho^{k}$. In other words, Fig. \ref{fig:WassJointPDFSynthetic} can be viewed as a nonparametric absolute error curve between $\rho^{\rm{MC}}$ and $\varrho^{k}$. Each snapshot value in Fig. \ref{fig:WassJointPDFSynthetic} is computed by solving the well-known optimal transport linear program (see e.g., \cite[Sec. 4.2-4.3]{halder2014probabilistic}, \cite[Ch. 3]{peyre2019computational}) using the corresponding time-varying weighted point clouds.

@@

\bibitem{wang2011numerical}
K.~Wang, and M.L.~Crow, ``Numerical simulation of stochastic differential algebraic equations for power system transient stability with random loads", \emph{2011 IEEE Power \& Energy Society General Meeting (PESGM)}, pp. 1--8, 2011. doi: 10.1109/PES.2011.6039188

\bibitem{milano2013systematic}
F.~Milano, and R.~Z{\'a}rate-Mi{\~n}ano, ``A systematic method to model power systems as stochastic differential algebraic equations", \emph{IEEE Transactions on Power Systems}, Vol. 28, No. 4, pp. 4537--4544, 2013.

@@

\bibitem{geng2020reach}
S.~Geng, and I.A.~Hiskens, ``Reach-Set Estimation for DAE Systems under Uncertainty and Disturbances Using Trajectory Sensitivity and Logarithmic Norm", \emph{21st IFAC World Congress}, Vol. 53, No. 2, pp. 1955--1961, 2020.

@@

\bibitem{silverman1998density}
B.W.~Silverman, \emph{Density Estimation for Statistics and Data Analysis}, 1st ed., Chapman \& Hall/CRC, 1998.

@@

\bibitem{bhatia2019bures}
R.~Bhatia, T.~Jain, and Y.~Lim, ``On the Bures--Wasserstein Distance between Positive Definite Matrices", \emph{Expositiones Mathematicae}, Vol. 37, No. 2, pp. 165--191, 2019.

@@
